---
title: "GSE96058 Breast Cancer RNA-seq Dataset"
author: "Daniel Incicau"
date: "2025-07-18"
output:
  html_document: default
  pdf_document: default
params:
  seed: 3213124
---

## Setup and Package Installation

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=12, warning=FALSE, message=FALSE, dev='png', dpi=300)

SEED = params$seed
set.seed(SEED)

library(tidyverse)
library(data.table)
library(readr)
library(umap)
library(Rtsne)
library(GEOquery)
library(ggplot2)
library(pheatmap)
library(patchwork)
library(cluster)
library(mclust) 
library(randomForest)
library(caret)
```


## Phase 1: Data Download & Preprocessing

### 1.1 Download Data from GEO

```{r}
gse <- getGEO("GSE96058", GSEMatrix = TRUE, AnnotGPL = TRUE)

# Load expression data
expression_file <- "GSE96058/GSE96058_gene_expression_3273_samples_and_136_replicates_transformed.csv.gz"
expression_data <- fread(expression_file, header = TRUE, check.names = FALSE)

# Expression Matrix
genes <- expression_data$V1
expression_matrix <- as.matrix(expression_data[, -1])
rownames(expression_matrix) <- genes

# Clinical Data
clinical_data_hiSeq <- pData(gse[[1]])
clinical_data_nextSeq <- pData(gse[[2]])
clinical_data <- rbind(clinical_data_hiSeq, clinical_data_nextSeq)
rownames(clinical_data) <- clinical_data[[1]]
clinical_data$sample_id <- rownames(clinical_data)
```

### 1.2 Prepare & Filter Data

```{r}
clean_clinical_data <- function(clinical_df) {
  colnames(clinical_df) <- make.names(colnames(clinical_df))
  colnames(clinical_df) <- gsub("\\.ch1", "", colnames(clinical_df))
  colnames(clinical_df) <- gsub("\\.+", "_", colnames(clinical_df))
  colnames(clinical_df) <- tolower(colnames(clinical_df))
  
  clinical_df_clean <- clinical_df %>%
    select(
      sample_id,
      age_at_diagnosis,
      tumor_size,
      lymph_node_status,
      er_status,
      pgr_status,
      her2_status,
      ki67_status,
      nhg,
      pam50_subtype,
      overall_survival_days,
      overall_survival_event
    )
  
  clinical_df_clean <- clinical_df_clean %>%
    mutate(across(everything(), ~na_if(., "NA")))
  
  clinical_df_clean <- clinical_df_clean %>%
    mutate(
      age_at_diagnosis = as.numeric(age_at_diagnosis),
      tumor_size = as.numeric(tumor_size),
      er_status = as.factor(er_status),
      pgr_status = as.factor(pgr_status),
      her2_status = as.factor(her2_status),
      ki67_status = as.factor(ki67_status),
      nhg = as.factor(nhg),
      pam50_subtype = as.factor(pam50_subtype),
      overall_survival_days = as.numeric(overall_survival_days),
      overall_survival_event = as.factor(overall_survival_event),
      lymph_node_status = as.factor(lymph_node_status)
    )
  
  return(clinical_df_clean)
}

filter_low_expression_genes <- function(
    expression_matrix,
    min_log2_expr = 1,          # Minimum log2 expression level
    min_sample_fraction = 0.1   # Fraction of samples that must express the gene
    ) {
  
  original_genes <- nrow(expression_matrix)
  min_samples <- ceiling(ncol(expression_matrix) * min_sample_fraction)
  samples_expressing <- rowSums(expression_matrix >= min_log2_expr)
  
  # Keep genes expressed in sufficient samples
  keep_expressed <- samples_expressing >= min_samples
  
  expression_filtered <- expression_matrix[keep_expressed, ]
  
  cat("Genes passing expression filter:", sum(keep_expressed), 
      "(", round(sum(keep_expressed)/original_genes*100, 1), "%)\n")
  cat("Genes removed:", sum(!keep_expressed), "\n\n")
  
  return(expression_filtered)
}

# Remove technical replicates (keep primary samples) 
primary_samples <- colnames(expression_matrix)[!grepl("repl", colnames(expression_matrix))]
expression_matrix <- expression_matrix[, primary_samples]

# Filter low expression genes
# log2(FPKM) > 0 means FPKM > 1
# Expressed in ≥10% of samples (≥341 samples)
expression_matrix <- filter_low_expression_genes(expression_matrix, min_log2_expr=0, min_sample_fraction=0.1)

# Clean clinical data
clinical_df <- clean_clinical_data(clinical_data)

# Remove not matching expression and clinical data
common_samples <- intersect(colnames(expression_matrix), rownames(clinical_df))
expression_matrix <- expression_matrix[, common_samples]
clinical_df <- clinical_df[common_samples, ]

# The resulting data was was post-processed by collapsing on 30,865 unique gene symbols (sum of FPKM values of each matching transcript), adding to each expression measurement 0.1 FPKM, and performing a log2 transformation.

cat("Dataset dimensions:", dim(expression_matrix), "\n")
cat("Number of genes:", nrow(expression_matrix), "\n")
cat("Number of samples:", ncol(expression_matrix), "\n")
cat("Clinical variables:", ncol(clinical_data), "\n")
cat("Expression range:", 
    round(min(expression_matrix, na.rm = TRUE), 2), "to", 
    round(max(expression_matrix, na.rm = TRUE), 2), "\n")

```

Check missing data from clinical.

```{r}
missing_summary <- clinical_df %>%
  select(
    age_at_diagnosis, tumor_size, lymph_node_status, er_status, 
    pgr_status, her2_status, ki67_status, nhg, pam50_subtype,
    overall_survival_days, overall_survival_event
  ) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Missing_Count") %>%
  mutate(Missing_Percentage = round(Missing_Count / nrow(clinical_df) * 100, 2)) %>%
  arrange(desc(Missing_Count))

print(missing_summary)
```

### 1.3 Quality Control Visualizations

```{r fig.width=12, fig.height=12}
set.seed(SEED)

# QC metrics calculation
qc_df <- data.frame(
  sample_id = colnames(expression_matrix),
  total_expression = colSums(expression_matrix),
  detected_genes = colSums(expression_matrix > 0),
  mean_expression = colMeans(expression_matrix),
  stringsAsFactors = FALSE
)

# Total expression per sample
p_total_expr <- ggplot(qc_df, aes(x = total_expression)) +
  geom_histogram(bins = 50, fill = "#3498db", alpha = 0.8) +
  geom_vline(xintercept = median(qc_df$total_expression), color = "red", linetype = "dashed") +
  labs(title = "Total Expression per Sample", x = "Sum of log2(FPKM + 0.1)", y = "Sample Count") +
  theme_minimal()

# Detected genes per sample
p_detected_genes <- ggplot(qc_df, aes(x = detected_genes)) +
  geom_histogram(bins = 50, fill = "#2ecc71", alpha = 0.8) +
  geom_vline(xintercept = median(qc_df$detected_genes), color = "red", linetype = "dashed") +
  labs(title = "Detected Genes per Sample", x = "# Genes with Expression > 0", y = "Sample Count") +
  theme_minimal()

# Sample-sample correlation matrix heatmap (subset)
sample_subset <- sample(colnames(expression_matrix), min(50, ncol(expression_matrix)))
cor_matrix <- cor(expression_matrix[, sample_subset])

pheatmap(cor_matrix,
         main = "Sample Correlation Matrix (Subset of 50)",
         show_rownames = FALSE, show_colnames = FALSE,
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "ward.D2",
         color = colorRampPalette(c("blue", "white", "red"))(100))

# Combine two QC histograms
p_total_expr + p_detected_genes
```

```{r fig.width=20, fig.height=12}
set.seed(SEED)

# Boxplot of expression distributions per sample (subset)
expr_subset <- expression_matrix[, sample(1:ncol(expression_matrix), min(10, ncol(expression_matrix)))]
expr_subset_df <- reshape2::melt(expr_subset)
colnames(expr_subset_df) <- c("Gene", "Sample", "Expression")

# Violin plot of expression per sample
p_violin <- ggplot(expr_subset_df, aes(x = Sample, y = Expression)) +
  geom_violin(fill = "lightblue", alpha = 0.7) +
  labs(title = "Expression Distribution", y = "log2(FPKM + 0.1)") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p_violin
```

## Stage 2: Data Exploration & Visualisation

### 2.1 Clinical Data

Explore age distribution, tumor size and PAM50 subtype

```{r fig.width=18, fig.height=6}
p_age <- ggplot(clinical_df, aes(x = age_at_diagnosis)) +
    geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
    geom_vline(xintercept = median(clinical_df$age_at_diagnosis, na.rm = TRUE), 
               color = "red", linetype = "dashed") +
    labs(title = "Age at Diagnosis Distribution", 
         x = "Age (years)", y = "Count") +
    theme_minimal()
  
p_tumor_size <- ggplot(clinical_df, aes(x = tumor_size)) +
    geom_histogram(bins = 30, fill = "darkgreen", alpha = 0.7) +
    geom_vline(xintercept = median(clinical_df$tumor_size, na.rm = TRUE), 
               color = "red", linetype = "dashed") +
    labs(title = "Tumor Size Distribution", 
         x = "Tumor Size (mm)", y = "Count") +
    theme_minimal()
  
p_pam50 <- ggplot(clinical_df, aes(x = pam50_subtype)) +
    geom_bar(fill = "purple", alpha = 0.7) +
    geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
    labs(title = "PAM50 Subtype Distribution", 
         x = "PAM50 Subtype", y = "Count") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
p_age + p_tumor_size + p_pam50
```

### 2.2 Expression Data

Compute low-dimensinal embeddings for expresion matrix (PCA, UMAP)

```{r}
set.seed(SEED)

# PCA
pca_result <- prcomp(t(expression_matrix), center = TRUE, scale. = TRUE)
pc_scores_df <- data.frame(pca_result$x[, 1:3], sample_id = rownames(pca_result$x))
pc_scores_df <- left_join(pc_scores_df, clinical_df, by = "sample_id")

# Variance explained
var_explained <- (pca_result$sdev^2 / sum(pca_result$sdev^2)) * 100
cumulative_variance <- cumsum(var_explained)

# UMAP
umap_result <- umap(t(expression_matrix))
umap_df <- data.frame(UMAP1 = umap_result$layout[,1],
                      UMAP2 = umap_result$layout[,2],
                      sample_id = colnames(expression_matrix))
umap_df <- left_join(umap_df, clinical_df, by = "sample_id")
```

Visualise low-dimensional embeddings by clinical subtype (PAM50)

```{r}
p_pca <- ggplot(pc_scores_df, aes(x = PC1, y = PC2, color = .data[["pam50_subtype"]])) +
    geom_point(alpha = 0.7, size = 2) +
    labs(title = "PCA of Gene Expression Data",
         x = paste0("PC1 (", round(var_explained[1], 2), "%)"),
         y = paste0("PC2 (", round(var_explained[2], 2), "%)")) +
    theme_minimal()

p_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = .data[["pam50_subtype"]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "UMAP of Gene Expression Data", color = "pam50_subtype") +
  theme_minimal()

p_pca
p_umap
```

Visualise low-dimensional embeddings by sample_id (check batch effects)

```{r}

p_pca_sample <- ggplot(pc_scores_df, aes(x = PC1, y = PC2, color = .data[["sample_id"]])) +
    geom_point(alpha = 0.7, size = 2) +
    labs(title = "PCA of Gene Expression Data",
         x = paste0("PC1 (", round(var_explained[1], 2), "%)"),
         y = paste0("PC2 (", round(var_explained[2], 2), "%)")) +
    theme_minimal() + 
    theme(legend.position = "none")


p_umap_sample <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = .data[["sample_id"]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "UMAP of Gene Expression Data", color = "sample_id") +
  theme_minimal() + 
  theme(legend.position = "none")

p_pca_sample
p_umap_sample
```

Visualize variance explained plots

```{r}
# Scree plot (top 20 PCs)
scree_data <- data.frame(
  PC = paste0("PC", 1:20),
  Variance = var_explained[1:20]
)

p_variance <- ggplot(scree_data, aes(x = reorder(PC, -Variance), y = Variance)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(Variance, 1)), vjust = -0.5, size = 3) +
  labs(title = "Varianice Explained", x = "Principal Component", y = "Variance Explained (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

cumsum_df <- data.frame(
  PC = 1:length(cumulative_variance),
  CumulativeVariance = cumulative_variance
)

p_variance_cumsum <- ggplot(cumsum_df, aes(x = PC, y = CumulativeVariance)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  geom_hline(yintercept = 90, linetype = "dashed", color = "red") +
  annotate("text", x = max(cumsum_df$PC) - 5, y = 91.5, label = "90% threshold", color = "red", size = 3) +
  labs(title = "Cumulative Variance Explained by PCs",
       x = "Number of Principal Components",
       y = "Cumulative Variance Explained (%)") +
  theme_minimal()

p_variance
p_variance_cumsum
```

## Stage 3: Analysis

### Stage 3.1: Feature Selection

Select only highly variable genes using the HVG method

```{r}
set.seed(SEED)

select_hvg_mean_var <- function(expression_matrix, top_n) {
  gene_means <- rowMeans(expression_matrix)
  gene_vars <- apply(expression_matrix, 1, var)
  
  # Fit loess curve to mean-variance relationship
  mv_fit <- loess(gene_vars ~ gene_means, span = 0.3)
  expected_var <- predict(mv_fit, gene_means)
  
  # Calculate standardized variance (observed/expected)
  var_ratio <- gene_vars / expected_var
  var_ratio[is.na(var_ratio)] <- 0
  
  # Select top genes by variance ratio
  top_hvg_genes <- names(sort(var_ratio, decreasing = TRUE))[1:top_n]
  return(list(genes = top_hvg_genes, var_ratio = var_ratio))
}

# Use HVG method (most robust for RNA-seq)
# top_n = 1000 (explains 90% of the variance)
hvg_result <- select_hvg_mean_var(expression_matrix, top_n = 1000)
top_hvg_genes <- hvg_result$genes
expression_selected <- expression_matrix[top_hvg_genes, ]

cat("\nSelected", length(top_hvg_genes), "highly variable genes for clustering out of", nrow(expression_matrix), "\n")
```

Visualize features selected

```{r fig.width=8, fig.height=4}
hvg_df <- data.frame(
  gene_id = names(hvg_result$var_ratio),
  mean_expr = rowMeans(expression_matrix),
  variance = apply(expression_matrix, 1, var),
  selected = names(hvg_result$var_ratio) %in% top_hvg_genes,
  stringsAsFactors = FALSE
)

p_hvg_selection <- ggplot(hvg_df, aes(x = mean_expr, y = variance, color = selected)) +
  geom_point(alpha = 0.6, size = 0.8) +
  scale_color_manual(values = c("FALSE" = "lightgray", "TRUE" = "red")) +
  geom_smooth(method = "loess", se = TRUE, color = "blue", linetype = "dashed") +
  labs(title = "Highly Variable Gene Selection",
       subtitle = paste("Selected", sum(hvg_df$selected), "genes (red points) out of", nrow(expression_matrix)),
       x = "Mean Expression", y = "Variance", color = "Selected") +
  theme_minimal()

p_hvg_selection
```

Compute low-dimensinal embeddings for selected expresion matrix (PCA, UMAP)

```{r}
set.seed(SEED)

# PCA
pca_result_selected <- prcomp(t(expression_selected), center = TRUE, scale. = TRUE)
pc_scores_selected_df <- data.frame(pca_result_selected$x[, 1:3], sample_id = rownames(pca_result_selected$x))
pc_scores_selected_df <- left_join(pc_scores_selected_df, clinical_df, by = "sample_id")


# UMAP
umap_result_selected <- umap(t(expression_selected))
umap_selected_df <- data.frame(UMAP1 = umap_result_selected$layout[,1],
                      UMAP2 = umap_result_selected$layout[,2],
                      sample_id = colnames(expression_selected))
umap_selected_df <- left_join(umap_selected_df, clinical_df, by = "sample_id")
```

Visualise low-dimensional embeddings by clinical subtype (PAM50)

```{r}
p_pca_selected <- ggplot(pc_scores_selected_df, aes(x = PC1, y = PC2, color = .data[["pam50_subtype"]])) +
    geom_point(alpha = 0.7, size = 2) +
    labs(title = "PCA of Gene Expression Data",
         x = paste0("PC1 (", round(var_explained[1], 2), "%)"),
         y = paste0("PC2 (", round(var_explained[2], 2), "%)")) +
    theme_minimal()

p_umap_selected <- ggplot(umap_selected_df, aes(x = UMAP1, y = UMAP2, color = .data[["pam50_subtype"]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "UMAP of Gene Expression Data", color = "pam50_subtype") +
  theme_minimal()

p_pca_selected
p_umap_selected
```

### Stage 3.2: Feature Normalization

Scale features using Z-score normalization

```{r}
processed <- t(scale(t(expression_selected)))
var_after_scale <- apply(processed, 1, var, na.rm = TRUE)
expression_scaled <- processed[var_after_scale > 0 & !is.na(var_after_scale), ]

cat("After preprocessing:", nrow(expression_scaled), "genes,", ncol(expression_scaled), "samples\n")
```

### Stage 3.3: Clustering

### 3.3.1 K-Means clustering

```{r}
set.seed(SEED)

ari_scores <- numeric()
cluster_assignments <- data.frame(
  sample_id = colnames(expression_scaled),
  true_label = clinical_df$pam50_subtype
)

# Run K-means for k = 3 to 7
for (k in 3:7) {
  cat("Run k-means clustering with k =", k)

  km <- kmeans(t(expression_scaled), centers = k, nstart = 25)
  cluster_assignments[[paste0("k=", k)]] <- as.factor(km$cluster)
  
  # Calculate ARI
  ari <- adjustedRandIndex(km$cluster, clinical_df$pam50_subtype)
  cat(". Result ARI =", ari, "\n")
  
  ari_scores[as.character(k)] <- ari
}

print("Adjusted Rand Index (ARI) for k-means clustering")
print(ari_scores)

write.csv(cluster_assignments, file = "kmeans_cluster_labels.csv", row.names = FALSE)
```

Visualise UMAP by clinical subtype (PAM50) / predicted K-Means labels

```{r fig.width=24, fig.height=12}
kmeans_umap_df <- left_join(umap_df, cluster_assignments, by = "sample_id")

p_umap_kmeans <- ggplot(kmeans_umap_df, aes(x = UMAP1, y = UMAP2, color = .data[["k=4"]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "UMAP of Gene Expression Data", color = "k=4") +
  theme_minimal()

p_umap + p_umap_kmeans
```

### 3.3.2 Hierarchical clustering

```{r}
set.seed(SEED)

# Perform hierarchical clustering
dist_matrix <- dist(t(expression_scaled), method = "euclidean")
hc <- hclust(dist_matrix, method = "ward.D2")

ari_scores <- numeric()
hclust_assignments <- data.frame(
  sample_id = colnames(expression_scaled),
  true_label = clinical_df$pam50_subtype
)

# Cut the tree at different levels of k (3 to 7)
for (k in 3:7) {
  cat("Run hierarchical clustering with k =", k)

  clusters <- cutree(hc, k = k)
  hclust_assignments[[paste0("k=", k)]] <- as.factor(clusters)
  
  # Calculate ARI
  ari <- adjustedRandIndex(clusters, clinical_df$pam50_subtype)
  cat(". Result ARI =", ari, "\n")
  
  ari_scores[as.character(k)] <- ari
}

# Print ARI scores
print("Adjusted Rand Index (ARI) for hierarchical clustering")
print(ari_scores)

# Save to CSV
write.csv(hclust_assignments, file = "hclust_cluster_labels.csv", row.names = FALSE)
```

Visualise UMAP by clinical subtype (PAM50) / predicted Hierarhical Clustering labels

```{r fig.width=24, fig.height=12}
hclust_umap_df <- left_join(umap_df, hclust_assignments, by = "sample_id")

p_umap_hclust <- ggplot(hclust_umap_df, aes(x = UMAP1, y = UMAP2, color = .data[["k=4"]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "UMAP of Gene Expression Data", color = "k=4") +
  theme_minimal()

p_umap + p_umap_hclust
```

### Stage 3.4: Supervised Analysis

### 3.4.1 Prepare Data for Classification

```{r}
# Check class distribution
cat("PAM50 Subtype Distribution:\n")
print(table(clinical_df$pam50_subtype))
cat("\nClass proportions:\n")
print(prop.table(table(clinical_df$pam50_subtype)))

prepared_data <- data.frame(
  pam50_subtype = clinical_df$pam50_subtype,
  t(expression_scaled)
)

cat("\nDataset for ML: ", nrow(prepared_data), "samples,", ncol(prepared_data)-1, "features.\n")
```

Since PAM50 subtypes are imbalanced, we need to handle that -> Undersampling

```{r}
set.seed(SEED)

# Random Undersampling
# Sample down to the size of the smallest class (Normal = 221)
min_class_size <- min(table(prepared_data$pam50_subtype))
cat("Minimum class size:", min_class_size, "\n")

# Undersample each class to match the smallest class
undersample_indices <- c()
for (class_name in names(table(prepared_data$pam50_subtype))) {
  class_indices <- which(prepared_data$pam50_subtype == class_name)
  sampled_indices <- sample(class_indices, min_class_size, replace = FALSE)
  undersample_indices <- c(undersample_indices, sampled_indices)
}

prepared_data_balanced <- prepared_data[undersample_indices, ]
print(table(prepared_data_balanced$pam50_subtype))
```
Cross validation setup

```{r}
set.seed(SEED)

evaluate_model <- function(predictions, actual) {
  cm <- confusionMatrix(predictions, actual)
  
  # Calculate per-class metrics
  per_class_metrics <- cm$byClass[, c("Sensitivity", "Specificity", "Precision", "F1")]
  
  # Overall metrics
  overall_acc <- cm$overall["Accuracy"]
  balanced_acc <- mean(cm$byClass[, "Balanced Accuracy"], na.rm = TRUE)
  
  return(list(
    confusion_matrix = cm$table,
    overall_accuracy = overall_acc,
    balanced_accuracy = balanced_acc,
    per_class_metrics = per_class_metrics,
    macro_f1 = mean(per_class_metrics[, "F1"], na.rm = TRUE)
  ))
}

summarize_cv_results <- function(fold_results) {
  summary_df <- data.frame()
  
  accuracies <- sapply(fold_results, function(x) x$overall_accuracy)
  balanced_accs <- sapply(fold_results, function(x) x$balanced_accuracy)
  macro_f1s <- sapply(fold_results, function(x) x$macro_f1)
    
  summary_df <- rbind(summary_df, data.frame(
    Mean_Accuracy = mean(accuracies, na.rm = TRUE),
    SD_Accuracy = sd(accuracies, na.rm = TRUE),
    Mean_Balanced_Accuracy = mean(balanced_accs, na.rm = TRUE),
    SD_Balanced_Accuracy = sd(balanced_accs, na.rm = TRUE),
    Mean_Macro_F1 = mean(macro_f1s, na.rm = TRUE),
    SD_Macro_F1 = sd(macro_f1s, na.rm = TRUE)
  ))
  
  return(summary_df)
}

cv_folds <- createFolds(prepared_data_balanced$pam50_subtype, k = 5, list = TRUE, returnTrain = FALSE)
```

### 3.4.2 Random Forest

```{r}
set.seed(SEED)

fold_results <- list()

for (fold_id in 1:length(cv_folds)) {
    cat("Running RandomForest CV for fold:", fold_id, "/", length(cv_folds), "\n")
  
    test_indices <- cv_folds[[fold_id]]
    train_indices <- setdiff(1:nrow(prepared_data_balanced), test_indices)
    
    train_data <- prepared_data_balanced[train_indices, ]
    test_data <- prepared_data_balanced[test_indices, ]
    
    rf_model <- randomForest(
        pam50_subtype ~ ., 
        data = train_data,
        ntree = 500,
        mtry = sqrt(ncol(train_data) - 1),
        importance = TRUE
    )
    
    predictions <- predict(rf_model, test_data)
    fold_results[[fold_id]] <- evaluate_model(predictions, test_data$pam50_subtype)
}

cv_summary <- summarize_cv_results(fold_results)
print(round(cv_summary, 4))
```
Train RandomForest model using 1st CV fold and extend predictions over the whole dataset.

```{r}
test_indices <- cv_folds[[1]]
train_indices <- setdiff(1:nrow(prepared_data_balanced), test_indices)
    
train_data <- prepared_data_balanced[train_indices, ]
    
final_model <- randomForest(
  pam50_subtype ~ ., 
  data = train_data,
  ntree = 500,
  mtry = sqrt(ncol(train_data) - 1),
  importance = TRUE
)

# Make predictions on full dataset using final model
test_data <- prepared_data[setdiff(rownames(prepared_data), rownames(train_data)), ]
final_predictions <- predict(final_model, test_data[, -1])

final_cm <- confusionMatrix(final_predictions, test_data$pam50_subtype)

cat("Final Model Performance:\n")
cat("Overall Accuracy:", round(final_cm$overall["Accuracy"], 4), "\n")
cat("Balanced Accuracy:", round(mean(final_cm$byClass[, "Balanced Accuracy"], na.rm = TRUE), 4), "\n")

p_cm <- ggplot(as.data.frame(final_cm$table), aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 4) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = paste("Confusion Matrix"),
       x = "True PAM50 Subtype", y = "Predicted PAM50 Subtype") +
  theme_minimal()

p_cm
```
```{r}
importance_df <- data.frame(
  Gene = rownames(final_model$importance),
  Mean_Decrease_Accuracy = final_model$importance[, "MeanDecreaseAccuracy"],
  Mean_Decrease_Gini = final_model$importance[, "MeanDecreaseGini"]
)

# Sort by importance and get top 20 features
top_features <- importance_df[order(importance_df$Mean_Decrease_Accuracy, decreasing = TRUE), ][1:20, ]

write.csv(cv_summary, file = "pam50_rf_cv_results.csv", row.names = FALSE)
write.csv(top_features, file = "pam50_rf_top_features.csv", row.names = FALSE)

print(top_features)
```

Visualise UMAP by clinical subtype (PAM50) / predicted Random Forest labels

```{r fig.width=24, fig.height=12}
rf_predictions <- data.frame(
  pam50_predicted = final_predictions,
  sample_id = names(final_predictions)
)

rf_umap_df <- left_join(rf_predictions, umap_df, by = "sample_id")

p_umap_rf <- ggplot(rf_umap_df, aes(x = UMAP1, y = UMAP2, color = .data[["pam50_predicted"]])) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "UMAP of Gene Expression Data (RF)", color = "pam50_predicted") +
  theme_minimal()

p_umap + p_umap_rf
```

---

## Session Information

```{r session_info}
sessionInfo()
```